Kindly Bent to Free Us

### 2019-07-08 kindly/infer/infer.tex
Inference and constraints

In order to determine if a variable is used in an affine manner, we must track
its uses and the associated kinds. For instance, in the expression
$(x,x)$, $x$ is used twice. If $x$ is of type $\tau$, which is of kind $k$,
we must add the constraint $\Cleq{k}{\kun}$.
%
In order to infer such constraints, our inference judgement will not only
take an environment as parameter but also return an environment, which
we call ``usage environment'', and which summarize how variables and borrows
are used. Usage environments follow the exact same grammar
as normal environment. In order to distinguish them more easily,
we will general note them $\Sv$.

@@

We note $\inferW{\Sv}{(C,\unif)}{\bf{\E}}{\bf{e}}{\tau}$ when
$e$\ has type $\tau$ in $\E$ under the constraints $C$ and unifier $\unif$
with a usage environment $\Sv$. $\E$ and $e$ are the input parameters of our
inference algorithm.
The rules are shown in \cref{rules:typing}.

\TODO{Explain/Cut}

@@

@@

\subsubsection{Copying and Splitting}
\label{infer:split}

\begin{figure}[!h]
  \centering
  \begin{mathpar}
    \ruleIPair
  \end{mathpar}
  \caption{The {\sc Pair$_I$} rule}
  \label{rule:infer:pair}
\end{figure}

\subsubsection{Regions}
\label{infer:regions}

\begin{figure}[!h]
  \centering
  \begin{mathpar}
    \ruleIRegion
  \end{mathpar}
  \caption{The {\sc Region$_I$} rule}
  \label{rule:infer:region}
\end{figure}

@@

@@

\TODO{Present the full constraint system}

We now place ourselves in a more general context where kinds are either variables
or constants belonging to a total bounded lattice $(\mathcal L, \lk_\Lat)$ ie.,
a lattice which admits a total order and upper and lower bounds ($l^\top$ and $l^\bot$).
We note lattice elements $l$ and $\glb_i l_i$ (resp. $\lub_i l_i$)
the greatest lower bound (resp. least upper bound) in $\mathcal L$.
%
Let $\CL$ the set of constraints in this lattice. We consider the
following base entailment rules in this system:
\begin{center}
  \begin{mathpar}
    \inferrule{l \lk_\Lat l'}{\entail{}{\Cleq{l}{l'}}}
    \and
    \inferrule{}{\entail{}{\Cleq{k}{l^\top}}}
    \and
    \inferrule{}{\entail{}{\Cleq{l^\bot}{k}}}
  \end{mathpar}
\end{center}

The normalization procedure is noted $\normalize{C_0}{\unif_0}$ where
$C_0\in \CL$ is a set of constraints and $\unif_0$ is a substitution.
It returns a constraint $C \in \mathcal S$ in
solved form and a unifier $\unif$.
The main idea of the algorithm is to first remove all the type equalities
by using regular Herbrand unification. After that, we only have
a set of inequalities among kinds, which we can consider as a relation.
We can then saturate the relation,
unify all kinds that are in the same equivalence classes to obtain
a most general unifier on kind variables,
remove all existentially quantified variables and
then minimize back the relation and apply various
simplification rules to make the resulting type easier to understand to users.

@@

\TODO{Add an example}

Our system is an extension of HM(X). Our goal is prove soundness and completeness
of our inference algorithm with respect to our syntax-directed system, which
assures us principal type inference.
We prove two properties:
\begin{itemize}
\item Our constraint system respects all the properties required by HM(X)
  for principal type inference.
\item Our extended inference algorithm still satisfies the soundness
  and completeness properties of HM(X). This is achieved by extending
  the original proofs.
\end{itemize}

This section only states the theorems (and the necessary definitions).
Details and proofs are provided in \cref{appendix:constraints,appendix:infer}.

\subsubsection{Constraint system}

We requires three properties from our constraint solving algorithms: it should
compute normal firms, such normal forms should be unique, and the
system should be regular.

@@

\subsubsection{Inference algorithm}

The first theorem states that our inference algorithm is sound
with respect to the syntax directed type system.

>>>>

In order to determine if a variable is used in an affine manner, we must track
its uses and the associated kinds. For instance, in the expression
$(x,x)$, $x$ is used twice. If $x$ is of type $\tau$, which is of kind $k$,
we must add the constraint $\Cleq{k}{\kun}$.
%
In order to infer such constraints, our inference judgement will not only
take an environment as parameter but also return an environment, which
we call ``usage environment'', and which summarize how variables and borrows
are used. Usage environments follow the exact same grammar
as normal environment. In order to distinguish them more easily,
we note them $\Sv$.

@@

We note $\inferW{\Sv}{(C,\unif)}{\bf{\E}}{\bf{e}}{\tau}$ when
$e$\ has type $\tau$ in $\E$ under the constraints $C$ and unifier $\unif$
with a usage environment $\Sv$. $\E$ and $e$ are the input parameters of our
inference algorithm.
We revisit the syntax-directed rules presented in \cref{sdtyping} to highlight the
novelties
of our inference algorithm and the differences with the syntax-directed
system.

@@

In our syntax directed system presented in \cref{sdtyping}, we ensured
that linear variables are not discarded at the \emph{leafs}, through
the {\sc Var} rule. In the inference algorithm, we operate in the opposite
direction: we collect data from the leafs, and only enforce linearity
at \emph{binders}. This is reflected in the {\sc Var$_I$} and
{\sc Abs$_I$} rules presented in \cref{rule:infer:envs}.
Typing for variables is very similar to traditional (non-affine) ML
typing. The only difference is that we collect
the fact that $x$ was used with the scheme $\schm$ by returning
a usage environment $\Sv = \{ \bvar{x}{\schm} \}$.
%
This usage environment is in turn used at each binder to enforce proper
usage of linear variable via the $\Weaken$ property.
We demonstrate how this is achieved for lambda expressions
in the {\sc Abs$_I$} rule.
First, we typecheck the body of the lambda and obtain a usage
environment $\Sv_x$. Just like in the syntax directed,
we introduce the constraint
$\Cleq{\Sv\Sdel{x}}{\kvar}$ which properly accounts for captures in
the body of the lambda expression. We also introduce the constraint
$\Weaken_{\bvar{x}{\sigma}}(\Sv)$. If $x$ was used $e$, $x$ is present
in $\Sv$ and this constraint is empty. Otherwise, if $x$ is
\emph{not} present in $\Sv$, its type should not have a linear
kind, which is enforced by the constraints $\Cleq{\sigma}{\kaff_\infty}$.
The $\Weaken$ constraint is introduce at each binding point in the program
such as let-bindings and pattern matching.

Note again here the use of normalization on complex constraints to ensure
that the inference algorithm always return the simplest possible
constraints and unifiers.

@@

\subsubsection{Splitting and Regions}
\label{infer:split}
\label{infer:regions}

As noted previously, in our inference algorithm, we infer
usage environments which describes how variables are used. In
\cref{sdtyping}, we described the {\sc Pair} and {\sc Region} rules, and their
respective actions on the environment. We now see the inference version
of these rules in \cref{rule:infer:envrules}.
While the rules are quite similar to the original version, the usage
environment $\Sv$ is now an \emph{output} of the algorithm.
As such, we use the ``inference'' version of the relations on
environment,
$\bsplit{C}{\Sv}{\bf{\Sv_1}}{\bf{\Sv_2}}$ and $\bregion{C}{\bf{x}}{\Sv}{\bf\Sv'}$,
which returns the necessary constraints and
are total and non ambiguous for their input parameters ($\Sv_1, \Sv_2$ and $\Sv'$, respectively).
We then simply collect all the constraints and normalize them.

\begin{figure}[!h]
  \centering
  \begin{mathpar}
    \ruleIPair

    \ruleIRegion
  \end{mathpar}
  \caption{The {\sc Pair$_I$} and {\sc Region$_I$} rules}
  \label{rule:infer:envrules}
\end{figure}

\subsubsection{Generalization and constraints}

So far, constraints have only been composed of a list of kind inequalities.
The internal language of constraints used for inference is in fact
richer.
To illustrate this, we now show the inference rule for let-bindings
in \cref{rule:infer:let}.
This rule combines several ingredients previously seen:
since let expressions are binders, we use $\Weaken$ on the bound
identifier $x$. Since let-expressions contain
two subexpressions, so we use the environment splitting relation,
$\bsplit{C_s}{\Sv}{\Sv_1}{(\Sv_2 \Sdel{x})}$. We carefully remove the $x$ from
the left environment, since it is not available outside of the expression
$e_2$, and thus should not be part of the returned usage environment.

As per tradition in ML languages, generalization is performed
on let-bindings.
Following HM(X), we note $(C_\schm, \schm) = \generalize{C}{\E}{\tau}$
the pair of a constraint and a scheme resulting from
generalization. The definition is provided in \cref{rule:infer:let}.
The type scheme $\schm$ is created by quantifying over all the appropriate
free variables and the current constraints.
The generated constraint $C_\schm$ uses a new projection operator,
noted $\Cproj{\tvar}{D}$, which
allow the creation of local variables inside the constraints.
This allow to encapsulate all the quantified variables in the global constraints.
It also reflect the fact that there
should exists at least one solution for $C$ for the scheme to be valid.
\citet{DBLP:journals/tapos/OderskySW99} give a detailed account
on the projection operators in HM inference.

\begin{figure}[!h]
  \centering
  \begin{mathpar}
    \ruleILet

    {\generalize{C}{\E}{\tau} =
      (\Cproj{\Multi{\kvar},\Multi{\tvar}}{C},
      \forall \Multi{\kvar},\Multi{\tvar}.\qual{C}{\tau})

      \text{ where }

      \Multi{\kvar},\Multi{\tvar} = (\fv{\tau}\cup\fv{C})\setminus\fv{\E}
    }
  \end{mathpar}
  \caption{The {\sc Let$_I$} rule}
  \label{rule:infer:let}
\end{figure}

@@

\TODO{Present this if there is space}

@@

\newcommand\A{\mathcal A}
\newcommand\SC{\mathcal S}

Before presenting our constraint solving algorithms,
we place ourselves in a more general context where kinds are either variables
or constants belonging to a total bounded lattice $(\mathcal L, \lk_\Lat)$ ie.,
a lattice which admits a total order and upper and lower bounds ($l^\top$ and $l^\bot$).
We note lattice elements $l$ and $\glb_i l_i$ (resp. $\lub_i l_i$)
the greatest lower bound (resp. least upper bound) in $\mathcal L$.
%
Let $\CL$ the set of constraints in this lattice.
The full grammar of constraints is shown in \cref{grammar:constraint}.
Constraints are made of kind inequalities, conjunctions and
projections, as shown previously, along with type unification
constraints. Since types might contain kinds (for instance, on the arrows),
type unification is oriented and noted $\leq$.

As before, entailment is noted $\entail{C}{D}$,
where $D$ is a consequence of the constraints $C$.
We say that $C$ and $D$ are equivalent, noted $C \equivC D$,
when $\entail{C}{D}$ and $\entail{D}{C}$.
The base rules for entailment are given in \cref{rules:entail}.
The rules for classic Herbrand unification and
projection operators are omitted for brevity (see
\citep{DBLP:journals/tapos/OderskySW99} for
details).

\begin{figure}[!h]
  \centering
  \begin{minipage}{0.32\linewidth}
    \begin{align*}
      C &::= \Cleq{\tau_1}{\tau_2} \\
        &\mid \Cleq{k_1}{k_2} \tag{Inequalities} \\
        &\mid C_1 \Cand C_2  \tag{Conjunction}\\
        &\mid \Cproj{\tvar}{C} 
        \mid \Cproj{\kvar}{C} \tag{Projection} 
    \end{align*}
    \caption{The constraint language}
    \label{grammar:constraint}
  \end{minipage}\hfill
  \begin{minipage}{0.64\linewidth}
    \begin{mathpar}
      \inferrule{}{\entail{}{\Cleq{k}{l^\top}}}
      \and
      \inferrule{}{\entail{}{\Cleq{l^\bot}{k}}}
      \and
      \inferrule
      {}{ \entail{}{\Cleq{\kvar}{\kvar}} }
      \and
      \inferrule
      { \forall i,\ \entail{C}{\Ceq{\tau_i}{\tau_i}}\\
      }
      { \entail{C}{\Cleq{\tapp{t}{(\tau_i)}}{\tapp{t}{(\tau'_i)}}} }
      \and
      \inferrule{l \lk_\Lat l'}{\entail{}{\Cleq{l}{l'}}}
      \and
      \inferrule
      { \entail{C}{\Cleq{\tau'_1}{\tau_1}}\\
        \entail{C}{\Cleq{\tau_2}{\tau'_2}}\\
        \entail{C}{\Cleq{k}{k'}}
      }
      { \entail{C}{\Cleq{\tau_1\tarr{k}\tau_2}{\tau'_1\tarr{k'}\tau'_2}} }
    \end{mathpar}
    \caption{Base entailment rules -- $\entail{C}{D}$ }
    \label{rules:entail}
  \end{minipage}
\end{figure}

We define the set of solved formed
$\mathcal S$ as the set of constraints $\CL$ quantified by $\equivC$.
We will show later that such constraints are in fact only composed of
kind inequalities, and thus correspond to the syntactic constraints
used in type and kind schemes.
%
We now define our normalization procedure, noted $\normalize{C_0}{\unif_0}$ where
$C_0\in \CL$ is a set of constraints and $\unif_0$ is a substitution.
It returns a constraint $C \in \mathcal S$ in
solved form and a unifier $\unif$.
The main idea of the algorithm is to first remove all the type equalities
by using regular Herbrand unification. After that, we only have
a set of inequalities among kinds, which we can consider as a relation.
We can then saturate the relation,
unify all kinds that are in the same equivalence classes to obtain
a most general unifier on kind variables,
remove all existentially quantified variables and
then minimize back the relation and apply various
simplification rules to make the resulting type easier to understand to users.

@@

\TODO{Add example}

This algorithm is complete and computes principal normal forms.
It can be completed by a set of ``best-effort'' simplification
rules which help reduce the size of inferred signatures
\citep{DBLP:conf/popl/PottierS02,DBLP:conf/aplas/Simonet03}.
In particular, our implementation and all the signatures presented in
\cref{motivation} use a variance-based simplification
where all covariants variables are replaced by their upper bounds
and contra-variant variables are replaced by their lower bounds.

We now show that this algorithms is well-behaved with respect to entailment
and computes unique normal forms.

@@

We now show that our extended inference algorithm still satisfies the soundness
and completeness properties of HM(X). This is achieved by extending
the original proofs.
%
The first theorem states that our inference algorithm is sound
with respect to the syntax directed type system.
